### 2008-10-09 trunk/Papers/rmrs-08/introduction.tex
some new stuff

Representing semantics as a logical form that supports automated
inference and model construction is vital for deeper language
engineering tasks, such as dialogue systems.  Logical forms
can be obtained from hand-crafted deep grammars
\cite{butt:etal:1999,copestake:flickinger:2000}, but this
tends to lack robustness: they fail to cover all
words and linguistic constructions and by design do not handle
ill-formed phrases. 
There has thus been a trend recently towards systems for
robust wide-coverage semantic construction (e.g., 
\cite{bos:etal:2004,zettlemoyer:collins:2007}).     
% COMMENT: this issue about optional arguments is really too cryptic.
% Basically, there is no way for assessing by *any* learned language
% processor whether a single word form that appears in
% a number of different syntactic valencies is an example of lexical
% sense ambiguity, or they are all examples of the same sense being
% used in syntactically different ways (i.e., we have optional
% arguments).  Only hand-crafted grammars can tell you this.  So, for
% example, Bos et al (2004) would get bet(x,y) for "I bet 5 pounds",
% and bet(x,y,z) for "I bet you 5 pounds" and bet(x,p) for "I bet it's
% going to rain": I.e., the vocab of the semantic formalism is
% constructed automatically from NL word forms, and the predicate
% symbols in that automatically constructed vocab vary in (a) valency,
% and (b) the semantic dependencies represented by a given argument
% position (e.g., the second argument of "bet" in "I bet 5 pounds" is
% 5 pounds, and for "I bet it's raining" it's "it's raining".
%

But there are certain semantic phenomena that these robust
approaches don't capture reliably, including quantifier scope,
optional arguments, and long-distance dependencies (for instance,
\newcite{clark:etal:2004} report that the parser used by
\newcite{bos:etal:2004} yields 63\% accuracy on object
extraction; e.g., {\em the man that I met\dots}).  Forcing a robust
parser to make a decision about these phenomena can therefore be
error-prone.  Depending on the application, it may be preferable to
give the parser the option to leave a semantic decision open when it's not
sufficiently informed---i.e., to compute a partial semantic
representation and to complete it later, using information
extraneous to the parser.

In this paper, we focus on an approach to representing partial
semantic information that supports this strategy: Robust Minimal
Recursion Semantics (\rmrs, \cite{copestake:2007a,copestake:2007b}).
\rmrs\ is designed to support underspecification of lexical ambiguity,
scope, and predicate-argument structure.  It is an emerging standard
for representing partial semantic information, and has been applied in
several implemented systems.  For instance, \newcite{copestake:2003}
and \newcite{frank:2004} use it to specify semantic components to
shallow parsers ranging in depth from {\sc pos} taggers to chunk
parsers and intermediate parsers such as {\sc rasp}
\cite{briscoe:etal:2006}.  {\sc mrs} analyses
\cite{copestake:etal:2005} derived from deep grammars, such as the
English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
are special cases of {\sc rmrs}.

>>>>

Representing semantics as a logical form that supports automated
inference and model construction is vital for deeper language
engineering tasks, such as dialogue systems.  Logical forms
can be obtained from hand-crafted deep grammars
\cite{butt:etal:1999,copestake:flickinger:2000}, but this
tends to lack robustness: they fail to cover all
words and constructions and by design don't handle
ill-formed phrases.

There has thus been a trend recently towards
robust wide-coverage semantic construction (e.g., 
\cite{bos:etal:2004,zettlemoyer:collins:2007}).     
% COMMENT: this issue about optional arguments is really too cryptic.
% Basically, there is no way for assessing by *any* learned language
% processor whether a single word form that appears in
% a number of different syntactic valencies is an example of lexical
% sense ambiguity, or they are all examples of the same sense being
% used in syntactically different ways (i.e., we have optional
% arguments).  Only hand-crafted grammars can tell you this.  So, for
% example, Bos et al (2004) would get bet(x,y) for "I bet 5 pounds",
% and bet(x,y,z) for "I bet you 5 pounds" and bet(x,p) for "I bet it's
% going to rain": I.e., the vocab of the semantic formalism is
% constructed automatically from NL word forms, and the predicate
% symbols in that automatically constructed vocab vary in (a) valency,
% and (b) the semantic dependencies represented by a given argument
% position (e.g., the second argument of "bet" in "I bet 5 pounds" is
% 5 pounds, and for "I bet it's raining" it's "it's raining".
% 
But there are certain semantic phenomena that these robust
approaches don't capture reliably, including quantifier scope,
optional arguments, and long-distance dependencies (for instance,
\newcite{clark:etal:2004} report that the parser used by
\newcite{bos:etal:2004} yields 63\% accuracy on object
extraction; e.g., {\em the man that I met\dots}).  Forcing a robust
parser to make a decision about these phenomena can therefore be
error-prone.  Depending on the application, it may be preferable to
give the parser the option to leave a semantic decision open when it's not
sufficiently informed---i.e., to compute a partial semantic
representation and to complete it later, using information
extraneous to the parser.

In this paper, we focus on an approach to representing partial
semantic information that supports this strategy: Robust Minimal
Recursion Semantics (\rmrs, \cite{copestake:2007a,copestake:2007b}).
\rmrs\ is designed to support underspecification of lexical information,
scope, and predicate-argument structure.  It is an emerging standard
for representing partial semantic information, and has been applied in
several implemented systems.  For instance, \newcite{copestake:2003}
and \newcite{frank:2004} use it to specify semantic components to
shallow parsers ranging in depth from {\sc pos} taggers to chunk
parsers and intermediate parsers such as {\sc rasp}
\cite{briscoe:etal:2006}.  {\sc mrs} analyses
\cite{copestake:etal:2005} derived from deep grammars, such as the
English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
are special cases of {\sc rmrs}.

### 2008-10-09 trunk/Papers/rmrs-08/motivation.tex
some new stuff

It exhibits several kinds of ambiguity, including a quantifier scope
ambiguity and lexical
ambiguities---e.g., the nouns ``cat'' and ``dog'' have 8 and 7
WordNet senses respectively.  Simplifying slightly by ignoring tense
information, two of these readings are shown as logical forms below;
these can be represented as trees as shown in Fig.~\ref{fig:1}.
%  (but see Section~\ref{sec:extensions})

@@

This \rmrs\ expresses only that certain predications must be made in
the semantic representation -- but it doesn't say anything about
semantic scope, about most arguments of the
predicates (note that $\sempred{\_chase\_v}(e)$ doesn't say who chases
whom), or about the coindexation of variables ($\sempred{\_every\_q}$
binds the variable $x_1$, whereas $\sempred{\_cat\_n}$ speaks about
$x_3$), and maintains the lexical ambiguities.  Technically, it
consists of six \emph{elementary predications} ({\sc ep}s), one for
each word lemma in the sentence; each of them is prefixed by a
\emph{label} and an \emph{anchor}, which are essentially variables
that refer to nodes in the trees in Fig.~\ref{fig:1}.  We can say that
the two trees \emph{satisfy} this \rmrs\ because it is possible to map
the labels and anchors into nodes in each tree and variable names like
$x_1$ and $x_3$ into variable names in the tree in such a way that the
predications of the nodes that labels and anchors are mapped to are 
consistent with those in the {\sc ep}s of the {\sc rmrs}.  For
instance, the first tree satisfies the \rmrs\ because $l_1$ and $a_1$
can map to the root of that tree and $x_1$ to $x$, and the root label
$\sem{\_every\_q\_1}$ is consistent with the {\sc ep} predicate
$\sempred{\_every\_q}$.

@@

This introduces two new types of atoms.   $x_3=x_4$
expresses that the two variables $x_3$ and $x_4$ must map to the same
variable in any fully specific logical form; e.g., both to the
variable $x$ in Fig.~\ref{fig:1}.  Atoms of the form $\ARG_i(a,z)$ and
$\ARG_i(a,h)$ express that the $i$-th child (counting from 0) of the
node to which the anchor $a$ refers is the variable name that $z$
denotes (or the node that the {\em hole} $h$ denotes).  So unlike
earlier underspecification formalisms, \rmrs\ can specify the
predicate of an atom separately from its (scopal and non-scopal)
arguments; this is necessary for supporting parsers where information
about lexical subcategorisation is absent. If we also allow atoms of
the form $\ARG_{\{2,3\}}(a,x)$ to express uncertainty as to whether
$x$ is the second or third child of the anchor $a$, then \rmrs\ can
even specify the arguments to a predicate while underspecifying their
position.  This can useful for specifying arguments to \_give\_v when a
parser doesn't handle unbounded dependencies and is faced with
{\em Which bone did you give the dog?} vs.\ {\em To which dog did you
  give the bone?}

@@

Finally, (\ref{ex:cat-erg}) uses predicate symbols like
$\sempred{\_dog\_n\_1}$ that are meant to be more specific than the
symbols like $\sempred{\_dog\_n}$ which the earlier \rmrs s used.
This reflects the fact that the deep grammar will perform some lexical
disambiguation, whereas the chunker and {\sc pos} tagger didn't.  The fact
that the former symbol should be more specific than the latter can be
represented using SPEC atoms like $\sempred{\_dog\_n\_1}\sqsubseteq
\sempred{\_dog\_n}$.
Note that even a deep grammar will not fully disambiguate to
semantic predicate symbols, such as WordNet senses, and so
$\sempred{\_dog\_n\_1}$ can still be consistent with multiple symbols
like $\sem{\_dog\_n\_1}$ and $\sem{\_dog\_n\_2}$ in the semantic
representation.  However, unlike the symbols in the \rmrs\ output from a
{\sc pos} tagger, those from a deep grammar will always be consistent
with symbols of the {\em same} arity, because a deep grammar
fully determines lexical subcategorisation.

In summary, \rmrs\ is a formalism for partial semantic
representations.  It allows us to represent the semantic information
that can be extracted from a wide range of {\sc nlp} tools in a uniform way.
This is useful for hybrid systems which fall back to shallower
analyses if deeper parsing is unsuccessful, or which try to match
deeply parsed queries against shallow parses of large text corpora;
and in fact, \rmrs\ is gaining popularity as a practical interchange
format for exactly these purposes \cite{copestake:2003,frank:2004}.
However, \rmrs\ is still relatively ad-hoc in that its
formal semantics is
not defined; we don't know, formally, what an \rmrs\ \emph{means} in
terms of semantic representations like (\ref{ex:fat-cat-1}) and
(\ref{ex:fat-cat-2}), and this hinders our ability to design efficient
algorithms for processing \rmrs. The purpose of this paper is to lay
the groundwork for fixing this problem.

>>>>

It exhibits several kinds of ambiguity, including a quantifier scope
ambiguity and lexical
ambiguities---e.g., the nouns ``cat'' and ``dog'' have 8 and 7
WordNet senses respectively.  Simplifying slightly by ignoring tense
information, two of its readings are shown as logical forms below;
these can be represented as trees as shown in Fig.~\ref{fig:1}.
%  (but see Section~\ref{sec:extensions})

@@

This \rmrs\ expresses only that certain predications must be made in
the semantic representation---but it doesn't say anything about
semantic scope, about most arguments of the
predicates (note that $\sempred{\_chase\_v}(e)$ doesn't say who chases
whom), or about the coindexation of variables ($\sempred{\_every\_q}$
binds the variable $x_1$, whereas $\sempred{\_cat\_n}$ speaks about
$x_3$), and it maintains the lexical ambiguities.  Technically, it
consists of six \emph{elementary predications} ({\sc ep}s), one for
each word lemma in the sentence; each of them is prefixed by a
\emph{label} and an \emph{anchor}, which are essentially variables
that refer to nodes in the trees in Fig.~\ref{fig:1}.  We can say that
the two trees \emph{satisfy} this \rmrs\ because it is possible to map
the labels and anchors into nodes in each tree and variable names like
$x_1$ and $x_3$ into variable names in the tree in such a way that the
predications of the nodes that labels and anchors are mapped to are 
consistent with those in the {\sc ep}s of the {\sc rmrs}.  For
instance, the first tree satisfies the \rmrs\ because $l_1$ and $a_1$
can map to the root of that tree and $x_1$ to $x$, and the root label
$\sem{\_every\_q\_1}$ is consistent with the {\sc ep} predicate
$\sempred{\_every\_q}$.

@@

This introduces two new types of atoms.   $x_3=x_4$
means that $x_3$ and $x_4$ map to the same
variable in any fully specific logical form; e.g., both to the
variable $x$ in Fig.~\ref{fig:1}.   $\ARG_i(a,z)$ and
$\ARG_i(a,h)$ express that the $i$-th child (counting from 0) of the
node to which the anchor $a$ refers is the variable name that $z$
denotes (or the node that the {\em hole} $h$ denotes).  So unlike
earlier underspecification formalisms, \rmrs\ can specify the
predicate of an atom separately from its
arguments; this is necessary for supporting parsers where information
about lexical subcategorisation is absent. If we also allow atoms of
the form $\ARG_{\{2,3\}}(a,x)$ to express uncertainty as to whether
$x$ is the second or third child of the anchor $a$, then \rmrs\ can
even specify the arguments to a predicate while underspecifying their
position.  This can useful for specifying arguments to \_give\_v when a
parser doesn't handle unbounded dependencies and is faced with
{\em Which bone did you give the dog?} vs.\ {\em To which dog did you
  give the bone?}

@@

Finally, (\ref{ex:cat-erg}) uses predicate symbols like
$\sempred{\_dog\_n\_1}$ that are meant to be more specific than the
symbols like $\sempred{\_dog\_n}$ which the earlier \rmrs s used.
This reflects the fact that the deep grammar will perform some lexical
disambiguation that the chunker and {\sc pos} tagger didn't.  The fact
that the former symbol should be more specific than the latter can be
represented using SPEC atoms like $\sempred{\_dog\_n\_1}\sqsubseteq
\sempred{\_dog\_n}$.
Note that even a deep grammar will not fully disambiguate to
semantic predicate symbols, such as WordNet senses, and so
$\sempred{\_dog\_n\_1}$ can still be consistent with multiple symbols
like $\sem{\_dog\_n\_1}$ and $\sem{\_dog\_n\_2}$ in the semantic
representation.  However, unlike the output of a
{\sc pos} tagger, an \rmrs\ symbol that's output by a deep grammar is
consistent 
with symbols that all have the {\em same} arity, because a deep grammar
fully determines lexical subcategorisation.

In summary, \rmrs\ allows us to represent the (partial) semantic information
that can be extracted from a wide range of {\sc nlp} tools in a uniform way.
This is useful for hybrid systems which fall back to shallower
analyses if deeper parsing is unsuccessful, or which try to match
deeply parsed queries against shallow parses of large text corpora;
and in fact, \rmrs\ is gaining popularity as a practical interchange
format for exactly these purposes \cite{copestake:2003,frank:2004}.
However, \rmrs\ is still relatively ad-hoc in that its
formal semantics is
not defined; we don't know, formally, what an \rmrs\ \emph{means} in
terms of semantic representations like (\ref{ex:fat-cat-1}) and
(\ref{ex:fat-cat-2}), and this hinders our ability to design efficient
algorithms for processing \rmrs. The purpose of this paper is to lay
the groundwork for fixing this problem.

### 2008-10-09 trunk/Papers/rmrs-08/rmrs.tex
some new stuff

\rmrs\ as presented here combines similarities to earlier
presentations of \rmrs\ \cite{copestake:2003,copestake:2007b} and to
{\sc clls}/dominance constraints \cite{egg:etal:2001}.  For the most part,
our syntax generalises that of older versions of \rmrs: We use
$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$ and
$\ARG_\N$ instead of $\ARG_n$, and the{\sc ep} $\ep{l}{a}{P(v)}$
(as in Section~\ref{sec:motivation}) is an abbreviation of
$\{\ep{l}{a}{P}, \ARG_{\{0\}}(a,v)\}$.  Similarly, we don't assume
that labels, anchors, and holes are syntactically different objects;
they receive their function from their positions in the formula.  One
major difference is that we use dominance rather than qeq; see
Section~\ref{sec:extensions} for a discussion.  Compared to dominance
constraints, the primary difference is that we now have a mechanism
for representing lexical ambiguity, and we can specify a predicate and
its arguments separately.

@@

\begin{definition}\label{defn:dominance}
  $u \dom v$ iff $u$ is a prefix of $v$, i.e.\ the node $u$ is above
  the node $v$ in the tree.  (This is the dominance relation familiar
  from \newcite{egg:etal:2001}.)

  $u \wedgedom v$ iff $u \dom v$, and all symbols on the path from $u$
  to $v$ (including $u$ but not $v$) are $\wedge$.
\end{definition}

The {\em satisfaction} relation between an \rmrs\ $\varphi$ and a
finite constructor tree $\tau$ is defined in terms of several
assignment functions.  First, we assume a node variable assignment
function $\alpha: \nvar \rightarrow D(\tau)$ that maps the node
variables in an \rmrs\ to the nodes of $\tau$.  Second, we assume a
base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
maps the base variables to nullary constructors
representing variables in the base language. Finally, we assume a
relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$ that maps
each \rmrs\ predicate symbol to a set of constructors from $\Sigma$.
As we'll see shortly, this relation allows us to underspecify lexical
ambiguities.

@@

In the context of robust semantic processing, this is a desirable
feature of \rmrs, because it means that when we enrich an \rmrs\
obtained from a shallow processor with more semantic
information---such as the relation symbols introduced by
syntactic 
constructions such as appositives, noun-noun compounds and free
adjuncts---we don't change the set of models;
we only \emph{restrict} the set of models further and further towards
the semantic representation we are trying to reconstruct.
Furthermore, it has been shown in the literature that a
dominance-constraint style semantics for underspecified
representations gives us more room to maneuver when developing
efficient solvers than an \mrs-style semantics
\cite{Althaus_etal:JoA}.

However, enumerating an infinite number of models is of course
infeasible.  For this reason, we will now transfer the concept of
\emph{solved forms} from dominance constraints to \rmrs.  An \rmrs\ in
solved form is guaranteed to be satisfiable, and thus each solved form
represents an infinite class of models.  However, each satisfiable
\rmrs\ has only a finite number of solved forms which partition the
space of possible models into classes with only `relevant'
differences as in Fig.~\ref{fig:1}.  A solver can then enumerate the
solved forms 
rather than all models.

@@

Let's look at an example for all these definitions.  All the \rmrs s
presented in Section~\ref{sec:motivation} (replacing $\qeq$ by
$\dom$), are in solved form; this is least obvious for
(\ref{ex:cat-erg}), but becomes clear once we notice that no label is
on the right-hand side of two dominance atoms.  However, this is
because (at least) one model of (\ref{ex:cat-erg}) is
shown in Fig.~\ref{fig:fat-black-cat}; this model
is problematic in several ways and in particular contains an unbound
variable $y$ even though it also contains a quantifier that binds $y$.
If we restrict the class of models to those in 
which such variables are bound (as \newcite{copestake:etal:2005} do),
we can add the information that the 
quantifiers outscope their bound variables to the \rmrs\ without
changing its models further---i.e., we add the atoms $h_3 \dom
l_5$ and $h_8 \dom l_5$.  Fig.~\ref{fig:fat-black-cat} is no longer a
model for the extended \rmrs, whic in turn is no longer in solved
form because the label $l_5$ is on the right-hand side of two
dominance edges.  Instead, it has the following two solved forms:

>>>>

Def.\ref{defn:rmrs-syntax}
combines similarities to earlier
presentations of \rmrs\ \cite{copestake:2003,copestake:2007b} and to
{\sc clls}/dominance constraints \cite{egg:etal:2001}.  For the most part,
our syntax generalises that of older versions of \rmrs: We use
$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$ and
$\ARG_\N$ instead of $\ARG_n$, and the {\sc ep} $\ep{l}{a}{P(v)}$
(as in Section~\ref{sec:motivation}) is an abbreviation of
$\{\ep{l}{a}{P}, \ARG_{\{0\}}(a,v)\}$.  Similarly, we don't assume
that labels, anchors, and holes are syntactically different objects;
they receive their function from their positions in the formula.  One
major difference is that we use dominance rather than qeq; see
Section~\ref{sec:extensions} for a discussion.  Compared to dominance
constraints, the primary difference is that we now have a mechanism
for representing lexical ambiguity, and we can specify a predicate and
its arguments separately.

@@

\begin{definition}\label{defn:dominance}
  $u \dom v$ iff $u$ is a prefix of $v$, i.e.\ the node $u$ is equal
  to or above
  the node $v$ in the tree.  (This is the dominance relation familiar
  from \newcite{egg:etal:2001}.)

  $u \wedgedom v$ iff $u \dom v$, and all symbols on the path from $u$
  to $v$ (not including $v$) are $\wedge$.
\end{definition}

The {\em satisfaction} relation between an \rmrs\ $\varphi$ and a
finite constructor tree $\tau$ is defined in terms of several
assignment functions.  First, a node variable assignment
function $\alpha: \nvar \rightarrow D(\tau)$ maps the node
variables in an \rmrs\ to the nodes of $\tau$.  Second, a
base language assignment function $g:\ovar \rightarrow \Sigma_0$ 
maps the base variables to nullary constructors
representing variables in the base language. Finally, a
relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$ maps
each \rmrs\ predicate symbol to a set of constructors from $\Sigma$.
As we'll see shortly, this relation allows an \rmrs\ to underspecify lexical
ambiguities.

@@

In the context of robust semantic processing, this is a desirable
feature of \rmrs, because it means that when we enrich an \rmrs\
obtained from a shallow processor with more semantic
information---such as the relation symbols introduced by
syntactic 
constructions such as appositives, noun-noun compounds and free
adjuncts---we don't change the set of models;
we only \emph{restrict} the set of models further and further towards
the semantic representation we are trying to reconstruct.
Furthermore, it has been shown in the literature that a
dominance-constraint style semantics for underspecified
representations gives us more room to manoeuvre when developing
efficient solvers than an \mrs-style semantics
\cite{Althaus_etal:JoA}.

However, enumerating an infinite number of models is of course
infeasible.  For this reason, we will now transfer the concept of
\emph{solved forms} from dominance constraints to \rmrs.  An \rmrs\ in
solved form is guaranteed to be satisfiable, and thus each solved form
represents an infinite class of models.  However, each satisfiable
\rmrs\ has only a finite number of solved forms which partition the
space of possible models into classes such that models within a class
differ only in `irrelevant' details.  A solver can then enumerate the
solved forms 
rather than all models.

@@

Let's look at an example for all these definitions.  All the \rmrs s
presented in Section~\ref{sec:motivation} (replacing $\qeq$ by
$\dom$), are in solved form; this is least obvious for
(\ref{ex:cat-erg}), but becomes clear once we notice that no label is
on the right-hand side of two dominance atoms.  However, this is
because (at least) one model of (\ref{ex:cat-erg}) is
shown in Fig.~\ref{fig:fat-black-cat}; this model
is problematic in several ways and in particular contains an unbound
variable $y$ even though it also contains a quantifier that binds $y$.
If we restrict the class of models to those in 
which such variables are bound (as \newcite{copestake:etal:2005} do),
we can add the information that the 
quantifiers outscope their bound variables to the \rmrs\ without
changing its models further---i.e., we add the atoms $h_3 \dom
l_5$ and $h_8 \dom l_5$.  Fig.~\ref{fig:fat-black-cat} is no longer a
model for the extended \rmrs, which in turn is no longer in solved
form because the label $l_5$ is on the right-hand side of two
dominance edges.  Instead, it has the following two solved forms:
